Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-19 paper.tex
More work on processing simulations.

\jkcomment{Make some sort of link here between this paper, which is
about msprime simulations, but that we can't really disentangle
simulations from their downstream processing.}
The point of simulations is to generate data and the efficiency with
which simulated data can be incorporated into an overall application
is an important issue.
For example, Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and more recent machine learning
methods~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable}
perform population genetic inference by using large numbers of
simulations as training data. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence data.
Modern approaches such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} can accommodate large
numbers of user-specified statistics that are individually weakly informative,
but jointly close to sufficient for the desired model or parameter.
However, the computational resources needed to calculate these statistics can be even
greater than those used to generate the simulated datasets.
Older software such as DIYABC~\citep{cornuet2008inferring}
or PopABC~\citep{lopes2009popabc} avoid this bottleneck by
developing their own specialised simulators, tightly integrating them
with the inference method. While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error.
As ABC inference has become useful in an ever-widening range of applications,
these all-in-one methods risk falling behind if they do not allow users to explore
the models and statistics of most interest.

\jkcomment{Other potential stuff to cover}
\begin{itemize}

\item What are the output formats of simulators? Newick, or sequence data
    basically. What's the problem with this? We don't have mutations.
    Much more efficient to have the tree and mutations than the
    sequence data. Fundamentally, the output of coalescent simulations
    should be trees, otherwise we are throwing a lot of information
    away.

\item Lots of examples of papers where people have been forced to
    create a modified fork of some simulator so that they can get
    access to some internal details, not exposed by the simulators
    interface. Tree sequences provide a complete record of the
    ancestral process, so we can do things like compute IBD
    segments and introgressed tracts directly on the output. See
    also ARG section for recording more information, though.

\item Side note somewhere: with efficient access to branch length
    statistics~\citep{ralph2020efficiently}, maybe we don't need to
    simulate mutations at all. (not in the ABC context, but elsewhere)

\item Tskit has facilited unprecedented interoperability between simulators:
    Forward simulations can also use tree
    sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
    3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
    tree sequences. Msprime can take tree sequences simulated by these forward
    simulators as input and complete the simulation (``recapitation'').
    See \cite{haller2018tree} for more details.
    (Although, this could be a discussion point?)

\end{itemize}

>>>>

\jkcomment{We need a link paragraph here between what this paper is
about (what we can simulate) and what the rest of this section is about
(efficiently using the \emph{results} of simulations). This is where
we should review the classical approaches. Then lead into the
the discussion of why we need simulations to be efficiently processable,
most importantly, in inference methods like ABC and ML.}

The abilily to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,schrider2018supervised,flagel2019unreasonable,
sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which the model and parameter space can be sampled from.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions of existing
simulators~\cite[e.g.][]{pavlidis2010msabc,quinto2018modeling},
tightly integrated with the inference method.
While this is certainly more efficient,
there is a great deal of duplication of effort and significant
room for error (and propagation of bugs).
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results even more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lacking a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output \jkcomment{Also there's the inherent limitation of
newick using branch lenghts; I think there's some citations for
the issues this causes. Probably in the original msprime paper.}.
This has been worked around by using approximate definitions of
concepts like identity by descent (cite carmi paper) \jkcomment{
any other computation methods developed just to work around the
crappiness of ms output?}
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

\jkcomment{Fit this in somewhere? Or maybe it should be a discussion point?}
Tskit has facilited unprecedented interoperability between simulators:
Forward simulations can also use tree
sequences~\citep{kelleher2018efficient,haller2018tree}. In particular, SLiM
3.0~\citep{haller2019slim} and fwdpp~\citep{thornton2014cpp} can now output
tree sequences. Msprime can take tree sequences simulated by these forward
simulators as input and complete the simulation (``recapitation'').
See \cite{haller2018tree} for more details.
